---
title: "U.S. Air Quality Data"
author: "Valentina Gonzalez Bohorquez"
date: "5/4/2021"
output: pdf_document
---

```{r setup, include=FALSE, echo=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE) 
```

## Project Goal: 
#### To compare yearly pollutant parameter data for an specific county and state. This project compares PM2.5 max values for Wake County, NC for years 2017 to 2020. Futhermore, it displays the PM2.5 avareage, max, min and standard deviation for Bergen County, New Jersey and New York, NY. 


#### API Access:
email=lvg23@scarletmail.rutgers.edu
key=cobaltheron82

## Load the libraries. 
```{r}

library(tidyverse)
library(jsonlite)
library(stats)
library(ggplot2)
library(gridExtra)
```

## Get desired parameter code.
```{r}

parameters_link <- "https://aqs.epa.gov/data/api/list/parametersByClass?email=lvg23@scarletmail.rutgers.edu&key=cobaltheron82&pc=ALL"

parametercodesdataframe <- parameters_link %>% jsonlite::fromJSON() %>% as.data.frame()
finalparametercodes_df <- parametercodesdataframe[5:6]

parametercode <- function(x){
  return (finalparametercodes_df[finalparametercodes_df$Data.value_represented == x,][1,1])
}
parametercode('PM2.5 - Local Conditions')
```
##### Created a function that returns pollutant parameter code given the pollutant parameter name.

## Get desired state code. 
```{r}
state_codes <- "https://aqs.epa.gov/data/api/list/states?email=lvg23@scarletmail.rutgers.edu&key=cobaltheron82"
statecodes_df <- state_codes %>% jsonlite::fromJSON() %>% as.data.frame()
finalstatecodes_df <- statecodes_df[5:6]

statecode <- function(x){
  return (finalstatecodes_df[finalstatecodes_df$Data.value_represented == x,][1,1])
}
statecode('North Carolina')
```
#### Created a function that returns state code given the state name.

## Get desired county code.
```{r}
countycode <- function(statename, countyname){
  c_url1<- "https://aqs.epa.gov/data/api/list/countiesByState?email=lvg23@scarletmail.rutgers.edu&key=cobaltheron82&state="
  c_url2 <- statecode(statename)
  countyurl <- paste(c_url1, c_url2,sep = "")
  tempdf <- countyurl %>% jsonlite::fromJSON() %>% as.data.frame()
  tempdf <- tempdf[5:6]
  return (tempdf[tempdf$Data.value_represented == countyname,][1,1])
}
countycode('North Carolina', 'Wake')
```
#### Created a function that returns county code given the state and county name.

## Daily summary data by county.
```{r}
tablefunction <- function(statename, countyname, parametername, year1){
  urlpart1 <- "https://aqs.epa.gov/data/api/dailyData/byCounty?email=lvg23@scarletmail.rutgers.edu&key=cobaltheron82&param="
  urlpart2<- "&bdate="
  urlpart3 <- "&edate="
  urlpart4 <- "&state="
  urlpart5 <- "&county="
  stateinput <- statecode(statename)
  countyinput <- countycode(statename, countyname)
  parameterinput <- parametercode(parametername)
  bdateinput <- paste(year1, "0101", sep = "")
  edateinput <- paste(year1, "1231", sep = "")
  urlinput <- paste(urlpart1,parameterinput,urlpart2,bdateinput,urlpart3,edateinput,urlpart4, stateinput, urlpart5, countyinput, sep = "")
  table_df <- urlinput %>% jsonlite::fromJSON() %>% as.data.frame()
  table_df <- cbind(table_df[16:16], table_df[23:23])
  table_df<-table_df[complete.cases(table_df),]
  datelist<-unique(table_df$Data.date_local) 
  
  colnameslist<-colnames(table_df) 
  
  for (i in datelist) {
    if (nrow(table_df[table_df$Data.date_local == i,]) > 1)
    {
      maxvalue <- max(table_df[table_df$Data.date_local == i,]$Data.first_max_value, na.rm = TRUE)
      table_df <- table_df[table_df$Data.date_local != i,]
      tempdf <- data.frame(i, maxvalue)
      names(tempdf) <- c("Data.date_local","Data.first_max_value")
      table_df <- rbind(table_df, tempdf)
    }
  }
  
  
  table_df <- table_df[order(as.Date(table_df$Data.date_local, format="%Y-%m-%d")),]
  return(table_df)
}

tablefunction("North Carolina", "Wake", "PM2.5 - Local Conditions", 2019)
```
#### Created a function that given the state, county, parameter, and year, returned the date and max value for that parameter. 

## Created a function that given the state, county, parameter, and year1, year2, year3, and year4, it returned a dataframe that displayed the date and max parameter values.
```{r}
severalyears <- function(statename, countyname, parametername, year1, year2, year3, year4){
  df1 <- tablefunction(statename, countyname, parametername, year1)
  df2 <- tablefunction(statename, countyname, parametername, year2)
  df3 <- tablefunction(statename, countyname, parametername, year3)
  df4 <- tablefunction(statename, countyname, parametername, year4)
  all <- rbind(df1, df2, df3, df4)
  allyears <- data.frame(all)
  allyears
}
Severalyears <- as.data.frame(severalyears("North Carolina", "Wake", "PM2.5 - Local Conditions", 2017, 2018, 2019, 2020))
```

The data frame displayed the PM2.5 max values in Wake County, NC for years 2017 to 2020.

## Plotted "Severalyears" and utilized LOESS method to draw a trend line.
```{r}
k <- ggplot(Severalyears,
            aes(x = as.Date(Data.date_local),
                y = Data.first_max_value)) +
  geom_point() + 
  geom_smooth(color = 'red') +
  labs(x = "Days", y = "PM2.5 Max Value",
       title = "Wake, NC Max PM2.5 Values for 2017 to 2020")
k
```

Between 2017 and mid-2018, the PM2.5 max value seemed to be steady. However, by late 2018 this value started to drop before slightly increasing in 2019. 

## Created a dataframe that that displayed the date and max parameter values, in order to take a closer look between 2019 and 2020. 
```{r}
df_year1 <- as.data.frame(tablefunction("North Carolina", "Wake", "PM2.5 - Local Conditions", 2019))
df_year2 <- as.data.frame(tablefunction("North Carolina", "Wake", "PM2.5 - Local Conditions", 2020))

bothyearsdf <- rbind(df_year1, df_year2)
bothyearsdf
```

## Plotted "bothyearsdf" and utilized LOESS method to draw a trend line.
```{r}
w <- ggplot(bothyearsdf,
            aes(x = as.Date(Data.date_local),
                y = Data.first_max_value)) +
  geom_point() + 
  geom_smooth(color = 'red') +
  labs(x = "Days", y = "PM2.5 Max Value",
       title = "Wake, NC Max PM2.5 Values for 2019 and 2020")
w
```

Between the end of 2019 and mid-2020, the PM2.5 pollutant max value decreased. The result could have been due to the COVID-19 quarantine. However, after July 2020, the PM2.5 max value seemed to began increasing, since the strict lock-down measures were lifted. 

## Created a final function that given the state, county, parameter, and year1 and year2, it returned a table that displayed the year, average, minimum, maximum and standard deviation values for those particular years.

```{r}
finalfunction<- function(statename, countyname, parametername, year1, year2){
  
  df1 <- tablefunction(statename, countyname, parametername, year1)
  df2 <- tablefunction(statename, countyname, parametername, year2)
  
  yearcolumn <- c(year1, year2)
  avg1 <- mean(df1$Data.first_max_value)
  avg2 <- mean(df2$Data.first_max_value)
  averagecolumn <- c(avg1, avg2)
  
  min1 <- min(df1$Data.first_max_value)
  min2 <- min(df2$Data.first_max_value)
  mincolumn <- c(min1, min2)
  
  max1 <- max(df1$Data.first_max_value)
  max2 <- max(df2$Data.first_max_value)
  maxcolumn <- c(max1, max2)
  
  sd1 <- sd(df1$Data.first_max_value)
  sd2 <- sd(df2$Data.first_max_value)
  sdcolumn <- c(sd1, sd2)
  
  sdcolumn <- c(sd(df1$Data.first_max_value), sd(df2$Data.first_max_value))
  df <- data.frame(yearcolumn, averagecolumn, mincolumn, maxcolumn, sdcolumn)
  
  df
  gridExtra::grid.table(df)
}
```
#### Note: Since the data pertains to the Max PM2.5 (or any other parameter) value for each day throughout the year, the averagecolumn corresponds to the average of all these max values. Simmilarly, the mincolumn corresponds to the minimum value out of all these max values, and vise versa for the max and standard deviation (std) values.

## Comparing 2016 and 2020 PM2.5 levels for Bergen County, New Jersey.
```{r}
finalfunction("New Jersey", "Bergen", "PM2.5 - Local Conditions", 2016, 2020)
```

The average max pollution value per day increased in Bergen County for 2020.

## Comparing 2016 and 2020 PM2.5 levels for New York,NY.
```{r}
finalfunction("New York", "New York", "PM2.5 - Local Conditions", 2016, 2020 )
```

The average max pollution value per day decreased in New York county for 2020. The reason could have been due to Covid-19. As people fled the virus in the city, they moved to more rural areas in Bergen County. In addition to quarantine being a factor.

